---
title: "Analysis of Categorical Data - Assignment - Phase 1"
author: "Amal Joy (s3644794) & Nupura Sanjay Sawle (s3639703)"
date: "08 September 2018"
output:
  pdf_document: 
    fig_caption: yes
    number_sections: yes
    toc: no
    toc_depth: 3
  html_document:
    df_print: paged
    toc: no
    toc_depth: '3'
linkcolor: blue
references:
- author:
  - family: Xie
    given: YiHui
    issued:
    - year: 2018
    publisher: RMIT University
    title: Dynamic Documents with R and knitr
  id: knitr
subtitle: MATH 1298 Analysis of Categorical Data Project Phase I
documentclass: article
---

\newpage

\tableofcontents

\newpage

# Introduction \label{sec1}

The aim of this project is to build a customer churn model for a telecommunication company to predict the customers who are about to get churned so that they can implement different business strategies to retain those customers before they actually get churned. The tool that I am using for this analysis is R-studio. This project will be conducted in 2 different phase where I will be exploring different data analysis techniques to accurately predict the churned customers. Phase 1 of this project will include the detailed descriptive statistical analysis of the data by making use of various R packages to build relevant charts, graphs, and interactions etc. Data preprocessing will be done to clean and transform the data to suit the prediction model. This section will address the issues with the dataset like missing values, outliers etc. The second phase of this project is solely for the model building and will include appropriate statistical procedures, the test of independence etc.

In this phase 1 of the project, the business rules of the telecommunication company are Identified to have a deep knowledge in customer churn in the telecommunication industry. Studies show that 5% the customer retention result in the  25%  to 95% the overall profit of the companies (Forbes, Jerry Jao, 2015). The major initiators of churn may include the quality of service, tariffs, unsatisfactory  post-sales service etc. Early identification of the customers who have the more chance of falling into the category of churned customers may help the marketing team to device cost-effective ways to build retention offers to target only those customers. It is also important to keep in mind that the incorrect predictions may lead to marketing team spending more time, resources and effort in trying to retain those customers who are not going to leave the company even if they were not provided with those offers. This means a company losing discounts offered to them.

# Dataset source and description

The following packages are used in this report for data preparation and data modeling.
```{r warning=FALSE, message=FALSE}
library(dplyr)
library(knitr)
library(kableExtra)
library(mlr)
library(cowplot)
library(ggplot2)
library(corrplot)
```

The data was read into a data file named 'telcom_churn'. Null values are replaced with 'NA' while reading the file.

```{r}
telcom_churn <- read.csv(
  "/Users/amaljoy/Study/Categorcal Data/Assignment 1/Telco-Customer-Churn.csv", 
  header=T,na.strings=c("","NA")) # Reading the data

dim(telcom_churn) # dimensions of the dataset
```

The dataset consists of 21 variables and 7043 observations. Each row in the dataset is the attributes associated to a customer, each column contains customer’s attributes. The customer attributes are provided below:
* customerID (Unique customer identification)  
* gender (female, male)  
* SeniorCitizen (Whether the customer is a senior citizen or not (1, 0))  
* Partner (Whether the customer has a partner or not (Yes, No))  
* Dependents (Whether the customer has dependents or not (Yes, No))  
* tenure (Number of months the customer has stayed with the company)  
* PhoneService (Whether the customer has a phone service or not (Yes, No))  
* MultipleLines (Whether the customer has multiple lines r not (Yes, No, No phone service)  
* InternetService (Customer’s internet service provider (DSL, Fiber optic, No)  
* OnlineSecurity (Whether the customer has online security or not (Yes, No, No internet service)  
* OnlineBackup (Whether the customer has an online backup or not (Yes, No, No internet service)  
* DeviceProtection (Whether the customer has device protection or not (Yes, No, No internet service)  
* TechSupport (Whether the customer has tech support or not (Yes, No, No internet service)  
* streamingTV (Whether the customer has streaming TV or not (Yes, No, No internet service)  
* streamingMovies (Whether the customer has streaming movies or not (Yes, No, No internet service)  
* Contract (The contract term of the customer (Month-to-month, One year, Two year)  
* Paperless billing (Whether the customer has paperless billing or not (Yes, No))  
* PaymentMethod (The customer’s payment method (Electronic check, Mailed check, Bank transfer (automatic), Credit card (automatic)))  
* MonthlyCharges (The amount charged to the customer monthly - numeric)  
* TotalCharges (The total amount charged to the customer - numeric)  
* Churn ( Whether the customer churned or not (Yes or No))  

The attribute churn will be the target variable. Given below is the first 5 observations in the dataset.

```{r echo=FALSE}
# Displaying heads of the table
kable(head(telcom_churn[,1:8]), caption = "Head of the data") %>%
  kable_styling("striped", full_width = T,
                bootstrap_options = c("striped", "hover","condensed"),
                latex_options = "HOLD_position",position = "center",
                                      font_size = 9) %>%
  row_spec(0, angle = 0,font_size = 7,bold = T, background = "#FAE5D3") %>%
  kableExtra::column_spec(1:8, border_right = T)

kable(head(telcom_churn[,9:14])) %>%
  kable_styling("striped", full_width = T,
                bootstrap_options = c("striped", "hover","condensed"),
                latex_options = "HOLD_position",position = "center",
                font_size = 9) %>%
  row_spec(0, angle = 0,font_size = 7,bold = T, background = "#FAE5D3") %>%
  column_spec(1:6, border_right = T)

kable(head(telcom_churn[,15:21])) %>%
  kable_styling("striped", full_width = T,
                bootstrap_options = c("striped", "hover","condensed"),
                latex_options = "HOLD_position",position = "center",
                font_size = 9) %>%
  row_spec(0, angle = 0,font_size = 7,bold = T, background = "#FAE5D3") %>%
  column_spec(1:6, border_left = T)
```

## Summary Statistics

Summary of the dataset is given below. It shows the summary of each variable including the levels in case of factors; range and central tendency values in case of numerical values.
```{r}
# Summary of the data
summary(telcom_churn)
```

From the summary statistics, it is clear that 'customerID' is an unusable attribute as is it a factor and is different for each observation. The male and female proportions in the dataset are almost similar. The attribute showing whether the customer is a senior citizen or not is a binary variable. But it is represented as a numerical variable. Data is distributed almost equally between people who have partners and those who do not. People with dependents are under-represented in this dataset. There is data for 72 months as represented by the variable 'tenure'. The minimum value of 'tenure' is 0 showing those customers who have joined within a month of collecting the data. Customers who don't have a phone service are very less in the sample and are almost 1/10th of those who have a phone service.  The variable 'MultipleLines' has 3 variables; customers who has/doesn't have multiple lines, and those who don't have a phone service. There are customers who don't subscribe for an internet service and those people are grouped separately as 'No internet service' in attributes 'OnlineSecurity', 'OnlineBackup', 'DeviceProtection' 'TechSupport', 'StreamingTV', and 'StreamingMovies'. There are three types of contracts 1 year, 2 year and month to month contracts. There are more number of  month to month contracts than both 1 year and 2 year contracts combined. Only a few people have signed up for paperless billing. The monthly charge for the customers varies from \$18.25 to \$118.75. The minimum value any customer has paid in total to the company is \$18.8 and the maximum amount any customer has paid in total to the company is \$8684.8. There are 11 observations with 'NA' values in 'TotalCharges'. The dataset is highly imbalanced in case of the target variable.  There are nearly three times data representing the active customers than the churned customers. This may lead to a biased model creation in the next phase of the project. We may have to deal with this data imbalance according to the type of model we are creating in the next phase.  

\pagebreak

# Data Preprocessing

The structure of the data is represented using the below line of code. 
```{r}
str(telcom_churn) # Structure of the data
```
'SeniorCitizen' has to be converted to a factor. It is labeled as 'Yes', and 'No'.

```{r}
telcom_churn$SeniorCitizen= factor(telcom_churn$SeniorCitizen, c(0,1), labels = c('No','Yes'),
                                   ordered = is.ordered(telcom_churn))
```

There are null values in the dataset. It can be checked using the following code chunk. 

```{r}
table(is.na(telcom_churn)) # Check for Null Values.
colSums(is.na(telcom_churn)) #totalcharges column have 11 missing values
```

'TotalCharges' has 11 'NA' values. For getting a closer picture of this null values, we can create a table of those customers with missing values as below: 
```{r,warning=FALSE}
kable(
  select(telcom_churn,gender, SeniorCitizen, tenure, Contract, 
         PaymentMethod, MonthlyCharges, Churn ) %>%
        filter(is.na(telcom_churn$TotalCharges)), caption = "Observations with missing values" ) %>%
  
  kable_styling("striped", full_width = T,
                bootstrap_options = c("striped", "hover","condensed"),
                latex_options = "HOLD_position",position = "center",
                font_size = 9) %>%
  
  row_spec(0, angle = 0,font_size = 7,bold = T, background = "#FAE5D3") %>%
  column_spec(1:7, border_left = T)
```

The table shows that the customers with missing values are all active customers. Their tenure is shown as zero which means that they all have joined within one month of the data collection. All of them are new customers. Customers who have joined recently will give less insight into the churn tendency of a customer. We cannot consider the attributes of a recent customer to train the model. For the stability of the model and better accuracy, we are considering only those customers who are customers of telco for at least 1 year, i.e 12 months. 

```{r}
len1 <- nrow(telcom_churn)
cat("Number of observations before:", len1)

telcom_churn <- subset(telcom_churn,telcom_churn$tenure>12) # creating new subset

len2 <- nrow(telcom_churn)
cat("\nNumber of observations after:", len2)
```
There are 2186 Observations being removed according to this criteria. Now we have 4857 churned/active customers who are with telco for more than one year. Checking if this removed our null values as well. 

```{r}
table(is.na(telcom_churn)) # Check for Null Values.
```

Since the minimum tenure is 12 months and also to reduce the Curse of Dimensionality and the time required to train the algorithm, we choose to factorise the variable 'tenure' in to 5 different groups; 13-24 months, 25-36 months, 37-48 months, 49-60 months, and 61-72 months. 
```{r}
telcom_churn$tenure <- cut(telcom_churn$tenure,breaks=5,dig.lab=2,labels=2:6)
```


\pagebreak

# Data Exploration

```{r}
data_summary <- telcom_churn %>% group_by(Churn) %>%
summarise(Percent = n() / nrow(.) * 100) # Creating percentage values

ggplot(data = data_summary,
mapping = aes(x = "", y = Percent, fill = Churn)) + 
  geom_bar(width = 1, stat = "identity") + 
  scale_y_continuous(breaks = round(cumsum(rev(data_summary$Percent)), 1)) + 
  coord_polar("y", start = 0) + 
  geom_text(aes(label = paste0(round(Percent), "%")), 
            position = position_stack(vjust = 0.5)) + 
  labs(x = NULL, y = NULL, fill = "Churn", 
       title = "Churn Customer Percentage") + 
  guides(fill = guide_legend(reverse = TRUE)) + 
  theme_classic() + 
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        plot.title = element_text(hjust =0.5))

```

Above graph clearly shows that out of 4857 customers 17% customers churned. 
```{r}
# Analysing Gender variable
ggplot(telcom_churn, aes(gender, ..count..,fill = Churn)) + 
  geom_bar(stat="count", position = "dodge") +  
  scale_fill_brewer(palette="Paired") + 
  ggtitle("Gender Vs Churn Plot") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(label=..count..),
            stat="count",position=position_dodge(0.5),vjust=-0.2)

```

This visualization shows that male and female both have similar chances to churn. Which suggest that gender variable have less impact on predicting which customers are likely to leave Telco telecom service.
```{r}
# Analyse senior citizen status variable
ggplot(telcom_churn, aes(SeniorCitizen,fill = Churn)) + 
  geom_bar(stat="count", position = "dodge") +  
  scale_fill_brewer(palette="Paired") +
  ggtitle("Senior Citizen Vs Churn Plot") + 
  theme(plot.title = element_text(hjust = 0.5))+
  geom_text(aes(label=..count..),
            stat="count",position=position_dodge(0.5),vjust=-0.2)
```

Above visualization shows that the number of senior citizens is less so a number of senior citizens churned are also less. 

```{r}
# Analyse Partner variable

ggplot(telcom_churn, aes(Partner,fill = Churn)) + 
  geom_bar(stat="count", position = "dodge") +  
  scale_fill_brewer(palette="Paired") +
  ggtitle("Partner Vs Churn Plot") + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  geom_text(aes(label=..count..),
            stat="count",position=position_dodge(0.5),vjust=-0.2)
```

This visualization shows that customer having partner have the lesser tendency of leaving Telco Telecom Services.

```{r}
# Analyse Dependents variable

ggplot(telcom_churn, aes(Dependents,fill = Churn)) + 
  geom_bar(stat="count", position = "dodge") +  
  scale_fill_brewer(palette="Paired") +
  ggtitle("Dependents Vs Churn Plot") + 
  theme(plot.title = element_text(hjust = 0.5))+
  geom_text(aes(label=..count..),
            stat="count",position=position_dodge(0.5),vjust=-0.2)
```

Above graph shows indicates that the number of customers having dependents is less hence the proportion of churning is as well.

```{r}
# Analyse tenure variable

ggplot(telcom_churn, aes(tenure,fill = Churn)) + 
  geom_bar(stat="count", position = "dodge") +  
  scale_fill_brewer(palette="Paired") +
  ggtitle("Tenure Vs Churn Plot") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(label=..count..),
            stat="count",position=position_dodge(0.5),vjust=-0.2)

```

Above plot represents that tenure in the year and churning customers are inversely proportional. Which implies that older customers have a lesser tendency of churning.

```{r}

# Analyse PhoneService variable

ggplot(telcom_churn, aes(PhoneService,fill=Churn)) + 
  geom_bar(stat="count", position = "dodge") +  
  scale_fill_brewer(palette="Paired") +
  ggtitle("Phone Service Vs Churn Plot") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(label=..count..),
            stat="count",position=position_dodge(0.5),vjust=-0.2)

```
This plot shows that in the examined data set very few customers don't use phone services.
```{r}
# Analyse MultipleLines variable

ggplot(telcom_churn, aes(MultipleLines,fill=Churn)) + 
  geom_bar(stat="count", position = "dodge") +  
  scale_fill_brewer(palette="Paired") +
  ggtitle("Multiple Lines Vs Churn Plot") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(label=..count..),
            stat="count",position=position_dodge(0.5),vjust=-0.2)

```

Above plot clearly, suggest that the number of customers with multiple lines are higher and churning tendency is higher as well.

```{r}
# Analyse InternetService variable

ggplot(telcom_churn, aes(InternetService,fill=Churn)) + 
  geom_bar(stat="count", position = "dodge") +  
  scale_fill_brewer(palette="Paired") +
  ggtitle("Internet Service Vs Churn Plot") + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  geom_text(aes(label=..count..),
            stat="count",position=position_dodge(0.5),vjust=-0.2)
```
Above graph shows that number of customers using DSL internet service and Fiber Optic internet service are almost same. But customers using Fiber Optic internet service are more likely to leave Telco Telcom Service.

```{r}
# Analyse OnlineSecurity variable

ggplot(telcom_churn, aes(OnlineSecurity,fill=Churn)) + 
  geom_bar(stat="count", position = "dodge") +  
  scale_fill_brewer(palette="Paired") +
  ggtitle("Online Security Vs Churn Plot") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(label=..count..),
            stat="count",position=position_dodge(0.5),vjust=-0.2)
```

This Visualization shows that customers with no online security services are more likely to leave.

```{r}
# Analyse OnlineBackup variable

ggplot(telcom_churn, aes(OnlineBackup,fill=Churn)) + 
  geom_bar(stat="count", position = "dodge") +  
  scale_fill_brewer(palette="Paired") +
  ggtitle("Online Backup Vs Churn Plot")+ 
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(label=..count..),
            stat="count",position=position_dodge(0.5),vjust=-0.2)
```

Above graph shows that the number of customers having an online backup is higher and customers without online backup are more likely to churn.

```{r}
# Analyse DeviceProtection variable

ggplot(telcom_churn, aes(DeviceProtection,fill=Churn)) + 
  geom_bar(stat="count", position = "dodge") +  
  scale_fill_brewer(palette="Paired") +
  ggtitle("Device Protection Vs Churn Plot") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(label=..count..),
            stat="count",position=position_dodge(0.5),vjust=-0.2)

```
Above visualization represents that the number of customers having Device Protection is higher and customers without Device Protection are more likely to churn.


```{r}
# Analyse TechSupport variable

ggplot(telcom_churn, aes(TechSupport, fill=Churn)) + 
  geom_bar(stat="count", position = "dodge") +  
  scale_fill_brewer(palette="Paired") +
  ggtitle("TechSupport Vs Churn Plot") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(label=..count..),
            stat="count",position=position_dodge(0.5),vjust=-0.2)

```
This graph clearly shows that the number of customers with tech support is higher and customers without online Tech support are more likely to churn.


```{r}
# Analyse StreamingTV variable

ggplot(telcom_churn, aes(StreamingTV,fill=Churn)) +
  geom_bar(stat="count", position = "dodge") +  
  scale_fill_brewer(palette="Paired") +
  ggtitle("TV Streaming Vs Churn Plot")+ 
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(label=..count..),
            stat="count",position=position_dodge(0.5),vjust=-0.2)

```

This graph clearly shows that the number of customers with TV Streaming services are higher so the number of customers to churn is also higher.

```{r}
# Analyse Streaming Movies variable

ggplot(telcom_churn, aes(StreamingMovies, fill=Churn)) + 
  geom_bar(stat="count", position = "dodge") +  
  scale_fill_brewer(palette="Paired") +
  ggtitle("Streaming Movies Vs Churn Plot") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(label=..count..),
            stat="count",position=position_dodge(0.5),vjust=-0.2)

```
This graph represents that the number of customers with Movie Streaming services are higher so the number of customers to churn is also higher.

```{r}
# Analyse Contract variable

ggplot(telcom_churn, aes(Contract, fill=Churn)) + 
  geom_bar(stat="count", position = "dodge") +  
  scale_fill_brewer(palette="Paired") +
  ggtitle("Contract Vs Churn Plot") + 
  theme(plot.title = element_text(hjust = 0.5))+
  geom_text(aes(label=..count..),
            stat="count",position=position_dodge(0.5),vjust=-0.2)

```
Above graph shows that the number of customers with two years of the contract are higher whereas customer with the month to month contract has a higher chance of leaving.

```{r}
# Analyse PaperlessBilling variable

ggplot(telcom_churn, aes(PaperlessBilling, fill=Churn)) + 
  geom_bar(stat="count", position = "dodge") +  
  scale_fill_brewer(palette="Paired") +
  ggtitle("Paperless Billing Vs Churn Plot") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(label=..count..),
            stat="count",position=position_dodge(0.5),vjust=-0.2)

```
This graph shows that the number of customers with Paperless Billing are higher so the proportion of churning is higher as well.

```{r}
# Analyse PaymentMethod variable

ggplot(telcom_churn, aes(PaymentMethod, fill=Churn)) + 
  geom_bar(stat="count", position = "dodge") +  
  scale_fill_brewer(palette="Paired") +
  ggtitle("Payment Method Vs Churn Plot") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(label=..count..),
            stat="count",position=position_dodge(0.5),vjust=-0.2) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

```
This visualization shows that majority of customers prefers Bank Transfer or Credit Card payment method and customers using electronic check payment method are more likely to churn. 

```{r}
# Analyse MonthlyCharges variable
ggplot(telcom_churn, aes(x=MonthlyCharges)) +
  geom_histogram(aes(y=..density..,fill=Churn),binwidth = 4) +
  geom_density(alpha=.05)+
  facet_grid(. ~ Churn)+ggtitle("Monthly Charges Vs Churn Plot") + 
  theme(plot.title = element_text(hjust = 0.5))
```
This plot shows that monthly charges are not monthly distributed for churned or not churned customers. Customers paying higher monthly charges are more so number of customers churning are more as well.


```{r}
# Analyse TotalCharges variable
ggplot(telcom_churn, aes(x=TotalCharges)) +
  geom_histogram(aes(y=..density..,fill=Churn),binwidth = 250) +
  geom_density(alpha=.05)+
  facet_grid(. ~ Churn)+
  ggtitle("Total Charges Vs Churn Plot")+ 
  theme(plot.title = element_text(hjust = 0.5))
```

This graph clearly shows that total charges are right-skewed for churned or not churned customers.

```{r}

ggplot(telcom_churn, aes(MonthlyCharges, TotalCharges, color = tenure)) +
  geom_point() + 
  scale_colour_brewer()+theme_gray()+
  ggtitle("Total Charges Vs Monthly Charges Across Tenure")+ 
  theme(plot.title = element_text(hjust = 0.5))
```
Above plot indicates that there is a strong positive correlation between monthly charges and total charges across all the Tenure levels. 
\pagebreak

# Results and Discussion
The data is highly imbalanced. In the dataset, there are only 17% of the customers who are churned. It is seen that gender doesn't have any effect on the deciding whether a customer is churned or not. As the tenure of the customer increases, the probability of the customer getting churned decreases. Out of all those people who don't use internet service at all, only 2% of the people got churned. This variable will have a strong relation effect in the model. Even though there is the highest number of people who take 2-year contracts and only 3% of them got churned. This is also an important variable in deciding if a customer gets churned or not. The proportion of churned customer is more in case of a month to month contract customers. Almost half of them got churned in the given dataset. People who are using the electronic check as the payment option has more chance of getting churned than people using any other payment options. We have seen from the multivariable scatter plot between total charges and monthly charges across the tenures that monthly charges are linearly and positively proportional to total charges across any particular tenure. So total charges may be a redundant variable and can create issues in the model creation unless removed. It is also interesting to note that as the tenure increases, the total charges and the monthly charges increases. This can mean that people may like to go for a higher payment plan over the years. 


# Conclusion
The data had several issues including missing values. Some of the factor variables were denoted as numerical and has to be converted into factors. Attributes were examined using graphs and numerical values. Several relations among variables were identified and some of the important variables were found. The data has been cleaned and processed for model creation.
